---
title: "edav_proj_crime_weather_traffic"
author: "sp_lee"
date: "Apr 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Bar charts of crime numbers against subway stations, in decreasing order**

```{r, echo=FALSE}
library(tidyverse)
subway_crime <- read_csv("crime/subway_crime_manhattan.csv")
crime_group_station <- subway_crime %>% group_by(Station_Name_w_Line) %>% summarize(crime_count = n())
crime_group_station_top <- subway_crime %>% group_by(Station_Name_w_Line) %>% summarize(crime_count = n()) %>% arrange(desc(crime_count)) %>% head(5)
crime_group_station_bottom <- subway_crime %>% group_by(Station_Name_w_Line) %>% summarize(crime_count = n()) %>% arrange(desc(crime_count)) %>% tail(10)

crime_group_station_crime_type <- subway_crime %>% group_by(Station_Name_w_Line, LAW_CAT_CD) %>% summarize(crime_count = n())

crime_group_station_crime_type <- crime_group_station_crime_type %>% arrange(desc(crime_count))
crime_group_station_felony <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "FELONY")
crime_group_station_misdemeanor <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "MISDEMEANOR")
crime_group_station_violation <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "VIOLATION")

crime_group_station_felony_top <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "FELONY") %>% arrange(desc(crime_count)) %>% head(5)
crime_group_station_misdemeanor_top <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "MISDEMEANOR")  %>% arrange(desc(crime_count)) %>% head(5)
crime_group_station_violation_top <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "VIOLATION") %>% arrange(desc(crime_count)) %>% head(5)

crime_group_station_felony_bottom <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "FELONY") %>% arrange(desc(crime_count)) %>% tail(10)
crime_group_station_misdemeanor_bottom <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "MISDEMEANOR")  %>% arrange(desc(crime_count)) %>% tail(10)
crime_group_station_violation_bottom <- crime_group_station_crime_type %>% filter(LAW_CAT_CD == "VIOLATION") %>% arrange(desc(crime_count)) %>% tail(10)

# Plot bar charts for the top stations (most crimes)

fig1.1 <- ggplot(crime_group_station_top, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + ggtitle('Top Crime Count by Manhattan Subway Stations') + coord_flip() + xlab("Subway Station") + ylab("Crime Count")
fig1.1

fig1.2 <- ggplot(crime_group_station_felony_top, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + coord_flip() + ggtitle('Top Crime Count (Felony) by Manhattan Subway Stations') + xlab("Subway Station") + ylab("Crime Count (Felony)")
fig1.2

fig1.3 <- ggplot(crime_group_station_misdemeanor_top, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + coord_flip() + ggtitle('Top Crime Count (Misdemeanor) by Manhattan Subway Stations') + xlab("Subway Station") + ylab("Crime Count (Misdemeanor)")
fig1.3

fig1.4 <- ggplot(crime_group_station_violation_top, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + coord_flip() + ggtitle('Top Crime Count (Violation) by Manhattan Subway Stations') + xlab("Subway Station") + ylab("Crime Count (Violation)")
fig1.4

# Plot bar charts for the bottom stations (least crimes)

fig1.5 <- ggplot(crime_group_station_bottom, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + ggtitle('Lowest Crime Count by Manhattan Subway Stations') + coord_flip() + xlab("Subway Station") + ylab("Crime Count")
fig1.5

fig1.6 <- ggplot(crime_group_station_felony_bottom, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + coord_flip() + ggtitle('Lowest Crime Count (Felony) by Manhattan Subway Stations') + xlab("Subway Station") + ylab("Crime Count (Felony)")
fig1.6

fig1.7 <- ggplot(crime_group_station_misdemeanor_bottom, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + coord_flip() + ggtitle('Lowest Crime Count (Misdemeanor) by Manhattan Subway Stations') + xlab("Subway Station") + ylab("Crime Count (Misdemeanor)")
fig1.7

fig1.8 <- ggplot(crime_group_station_violation_bottom, aes(x = reorder(Station_Name_w_Line, crime_count), y = crime_count)) + geom_col(fill = "pink") + coord_flip() + ggtitle('Lowest Crime Count (Violation) by Manhattan Subway Stations') + xlab("Subway Station") + ylab("Crime Count (Violation)")
fig1.8


```


By looking at these bar charts displaying the top 5 stations in terms of crime count, we see that the crime count by station for the 3 different crime types were closely related.  For example, 125 ST (line 4, 5, 6) and 23 ST were in the top 5 for all 3 types of crime.  We see that the similarity between misdemeanor and violation was strongest among the 3 possible pairings of crime type (i.e., felony-misdemeanor, felony-voilation, misdemeanor-violation).  In a way, this was not surprising because misdemeanor and violation are more similar than felony which is a more serious type of crime.

Out of interest, we also looked at the 10 "safest" stations.  The patterns are a little harder to infer compared with the most dangerous stations because there are many ties.  Some interesting observations include: wall street station is one of the safest, whether we are looking at felony or misdemeanor.  And Columbia on 116th street is also one of the safest for misdemeanor! 

**Box plots of crime numbers by time of the day**

```{r, echo=FALSE}
crime_group_station_time_interval <- subway_crime %>% group_by(Station_Name_w_Line, hour_interval) %>% summarize(crime_count = n())
crime_group_station_crime_type_time_interval <- subway_crime %>% group_by(Station_Name_w_Line, hour_interval, LAW_CAT_CD) %>% summarize(crime_count = n())

time_levels = c("0000-0400", "0400-0800", "0800-1200", "1200-1600", "1600-2000", "2000-0000")

crime_group_station_time_interval$hour_interval <- factor(crime_group_station_time_interval$hour_interval, levels = time_levels)

crime_group_station_crime_type_time_interval$hour_interval <- factor(crime_group_station_crime_type_time_interval$hour_interval, levels = time_levels)

crime_group_station_felony_time_interval <- crime_group_station_crime_type_time_interval%>% filter(LAW_CAT_CD == "FELONY")
crime_group_station_misdemeanor_time_interval <- crime_group_station_crime_type_time_interval%>% filter(LAW_CAT_CD == "MISDEMEANOR")
crime_group_station_violation_time_interval <- crime_group_station_crime_type_time_interval%>% filter(LAW_CAT_CD == "VIOLATION")

fig2.1 <- ggplot(crime_group_station_time_interval, aes(hour_interval, crime_count)) + geom_boxplot(outlier.colour="red") + coord_flip() + ggtitle('Crime Count by Time Period of the Day') + xlab("Time Interval") + ylab("Crime Count")
fig2.1
fig2.2 <- ggplot(crime_group_station_time_interval, aes(hour_interval, crime_count)) + geom_boxplot(outlier.colour="red") + coord_flip() + ggtitle('Crime Count by Time Period of the Day') + xlab("Time Interval") + ylab("Crime Count") + geom_text(label=crime_group_station_time_interval$Station_Name_w_Line, check_overlap = TRUE, nudge_x=0.2, nudge_y=-0.3) 
fig2.2
fig2.3 <- ggplot(crime_group_station_felony_time_interval, aes(hour_interval, crime_count)) + geom_boxplot(outlier.colour="red") + coord_flip() + ggtitle('Crime Count (Felony) by Time Period of the Day') + xlab("Time Interval") + ylab("Crime Count")
fig2.3
fig2.4 <- ggplot(crime_group_station_misdemeanor_time_interval, aes(hour_interval, crime_count)) + geom_boxplot(outlier.colour="red") + coord_flip() + ggtitle('Crime Count (Misdemeanor) by Time Period of the Day') + xlab("Time Interval") + ylab("Crime Count")
fig2.4
fig2.5 <- ggplot(crime_group_station_violation_time_interval, aes(hour_interval, crime_count)) + geom_boxplot(outlier.colour="red") + coord_flip() + ggtitle('Crime Count (Violation) by Time Period of the Day') + xlab("Time Interval") + ylab("Crime Count")
fig2.5
```


For overall crime count, the median was rather consistent across the time periods.  The number of crimes committed was highest from 1600-2000hr, which coincided with the evening peak period.  Based on that, we expected the morning peak (0800-1200hr) to display the next highest crime count, but the data did not support that.  Instead, 1200-1600hr showed the second highest crime count based on median.  Also, variance (as indicated by length of the box) was highest for the time periods with the highest median crime count. The patterns we saw in the overall count were similarly visible for felony and misdemeanor.  For violation, the only similarity with the other crime types was that 1600-200hr was the period with the highest crime rate.

**Scatter plots of Subway Human Traffic against Crime Count (by weekend, weekday; by crime type)**
*One data point = one day of the year*

```{r, echo=FALSE}
crime_count_date <- subway_crime %>% group_by(CMPLNT_FR_DT) %>% summarize(crime_count = n())
crime_type_count_date <- subway_crime %>% group_by(CMPLNT_FR_DT, LAW_CAT_CD) %>% summarize(crime_count = n())

subway_traffic <- read_csv("data/2015_manhattan_turnstile_usage2.csv")
subway_total_traffic <- subway_traffic %>% mutate(total_vol = entry_volume + exit_volume)
subway_total_traffic_wkday <- subway_total_traffic %>% filter(day_of_week %in% c(0, 1, 2, 3, 4) & is_holiday == "False")
subway_total_traffic_wkend <- subway_total_traffic %>% filter(day_of_week %in% c(5, 6) | is_holiday == "True")
subway_total_traffic_wkday_day <- subway_total_traffic_wkday %>% group_by(date) %>% summarize(traffic = sum(total_vol))
subway_total_traffic_wkend_day <- subway_total_traffic_wkend %>% group_by(date) %>% summarize(traffic = sum(total_vol))

crime_count_traffic_wkday <- inner_join(subway_total_traffic_wkday_day, crime_count_date, c("date" = "CMPLNT_FR_DT"))
crime_count_traffic_wkend <- inner_join(subway_total_traffic_wkend_day, crime_count_date, c("date" = "CMPLNT_FR_DT"))
crime_type_count_traffic_wkday <- inner_join(subway_total_traffic_wkday_day, crime_type_count_date, c("date" = "CMPLNT_FR_DT"))
crime_type_count_traffic_wkend <- inner_join(subway_total_traffic_wkend_day, crime_type_count_date, c("date" = "CMPLNT_FR_DT"))

felony_count_traffic_wkday <- crime_type_count_traffic_wkday %>% filter(LAW_CAT_CD == "FELONY")
misdemeanor_count_traffic_wkday <- crime_type_count_traffic_wkday %>% filter(LAW_CAT_CD == "MISDEMEANOR")
violation_count_traffic_wkday <- crime_type_count_traffic_wkday %>% filter(LAW_CAT_CD == "VIOLATION")
felony_count_traffic_wkend <- crime_type_count_traffic_wkend %>% filter(LAW_CAT_CD == "FELONY")
misdemeanor_count_traffic_wkend <- crime_type_count_traffic_wkend %>% filter(LAW_CAT_CD == "MISDEMEANOR")
violation_count_traffic_wkend <- crime_type_count_traffic_wkend %>% filter(LAW_CAT_CD == "VIOLATION")

fig3.1 <- ggplot(crime_count_traffic_wkday, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Crime Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.1
fig3.2 <- ggplot(crime_count_traffic_wkend, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekend Subway Human Traffic against Crime Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.2

fig3.3 <- ggplot(felony_count_traffic_wkday, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Felony Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.3
fig3.4 <- ggplot(felony_count_traffic_wkend, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekend Subway Human Traffic against Felony Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.4

fig3.5 <- ggplot(misdemeanor_count_traffic_wkday, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Misdemeanor Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.5
fig3.6 <- ggplot(misdemeanor_count_traffic_wkend, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Misdemeanor Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.6

fig3.7 <- ggplot(violation_count_traffic_wkday, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Violation Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.7
fig3.8 <- ggplot(violation_count_traffic_wkend, aes(x = crime_count, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Violation Count - 1 point = one day') + xlab("Crime Count") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig3.8

```


For the above and subsequent similar scatter plots, each data point on the graph represented a single day (e.g., total human traffic across all stations for that day, total crimes committed across all stations for that day).  

We created the above scatter plots to investigate if there was a relationship between human traffic and crime rate at subway stations.

We plotted the charts separately by weekday and weekend to isolate any effects that weekday vs weekend might have on the relationship between human traffic count and crime count.

Overall, there was a positive correlation between human traffic and crime count for both weekday and weekend.  A similar pattern was observed for felony and misdemeanor.  However, the relationship was less obvious for violation for which the sample sizes were small.

**Scatter plots of Subway Human Traffic against Rainfall (by weekend, weekday)**
*One data point = one day of the year*

```{r, echo=FALSE}
subway_total_traffic_wkday_day
subway_total_traffic_wkend_day
rainfall <- read_csv("weather/precipitation.csv")
rainfall_day <- rainfall %>% group_by(date) %>% summarize(total_rainfall = sum(total_precip))

rainfall_traffic_wkday <- inner_join(subway_total_traffic_wkday_day, rainfall_day)
rainfall_traffic_wkend <- inner_join(subway_total_traffic_wkend_day, rainfall_day)

rainfall_traffic_wkday
rainfall_traffic_wkend

fig4.1 <- ggplot(rainfall_traffic_wkday, aes(x = total_rainfall, y = traffic)) + geom_point(color='blue') + ggtitle('Weekday Subway Human Traffic against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig4.1
fig4.2 <- ggplot(rainfall_traffic_wkend, aes(x = total_rainfall, y = traffic)) + geom_point(color='blue') + ggtitle('Weekend Subway Human Traffic against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Human Traffic Count") + stat_smooth(method = "lm", col = "red")
fig4.2

```


The above scatter plots investigated if there was a relationship between rainfall and human traffic at subway stations.

For weekdays, we saw that human traffic was not much influenced by rainfall, which was not surprising because everyone had to go to work/school regardless of whether or not it was raining.   For weekends, we saw a stronger negative relationship between rainfall and human traffic, which made sense because people might cancel their outdoor activities or leisure travelling plans depending on the weather. 

We also saw that most of the data points were clustered around the y-axis, which was due to the fact that on most days there were no rain.

**Scatter plots of Crime Count against Rainfall (by crime type)**
*One data point = one day of the year*

```{r, echo=FALSE}
crime_count_date
crime_type_count_date

rainfall_day

rainfall_wkday <- rainfall_traffic_wkday %>% select(date, total_rainfall)
rainfall_wkend <- rainfall_traffic_wkend %>% select(date, total_rainfall)

crime_rainfall <- inner_join(rainfall_day, crime_count_date, c("date" = "CMPLNT_FR_DT"))
crime_rainfall

crime_type_count_date

felony_count_date <- crime_type_count_date %>% filter(LAW_CAT_CD == "FELONY")
misdemeanor_count_date <- crime_type_count_date %>% filter(LAW_CAT_CD == "MISDEMEANOR")
violation_count_date <- crime_type_count_date %>% filter(LAW_CAT_CD == "VIOLATION")

felony_rainfall <- inner_join(rainfall_day, felony_count_date, c("date" = "CMPLNT_FR_DT"))
misdemeanor_rainfall <- inner_join(rainfall_day, misdemeanor_count_date, c("date" = "CMPLNT_FR_DT"))
violation_rainfall <- inner_join(rainfall_day, violation_count_date, c("date" = "CMPLNT_FR_DT"))

crime_rainfall_wkend <- inner_join(rainfall_wkend, crime_count_date, c("date" = "CMPLNT_FR_DT"))

fig5.1 <- ggplot(crime_rainfall, aes(x = total_rainfall, y = crime_count)) + geom_point(color='blue') + ggtitle('Crime Count against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig5.1

fig5.2 <- ggplot(felony_rainfall, aes(x = total_rainfall, y = crime_count)) + geom_point(color='blue') + ggtitle('Felony Count against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig5.2

fig5.3 <- ggplot(misdemeanor_rainfall, aes(x = total_rainfall, y = crime_count)) + geom_point(color='blue') + ggtitle('Misdemeanor Count against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig5.3

fig5.4 <- ggplot(violation_rainfall, aes(x = total_rainfall, y = crime_count)) + geom_point(color='blue') + ggtitle('Violation Count against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig5.4

fig5.5 <- ggplot(crime_rainfall_wkend, aes(x = total_rainfall, y = crime_count)) + geom_point(color='blue') + ggtitle('Weekend Crime Count against Rainfall - 1 point = one day') + xlab("Rainfall") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig5.5

```


The above scatter plots investigated if there was a relationship between crime rate and rainfall at subway stations.

There did not seem to be a strong relationship between overall crime count and rainfall but we did notice that on days with heavy rain, crime count was never high, especially when we drilled down to look at misdemenor and violations.  One explanation would be that heavy rainfall might have deterred potential offenders from travelling to the subway stations.   

However, we noticed that this was not true for felony.  High counts of felony were observed even for days with heavy rain, suggesting that felony was less dependent on the weather.

Previously, we saw that subway human traffic was sensitive to rainfall on weekends but not so much on weekdays.  Therefore, we wondered if crime count would likewise be more sensitive to rainfall on weekends.  Based on visual inspection, it did not seem like weekend changed the relationship between crime rate and rainfall.  

**Scatter plots of Crime Count against Subway Human Traffic (by crime type; by weekend, weekday)**
*One data point = one subway station*

```{r, echo=FALSE}
wkday <- subway_total_traffic %>% filter(day_of_week %in% c(0, 1, 2, 3, 4) & is_holiday == "False") %>% select(date)
wkend <- subway_total_traffic %>% filter(day_of_week %in% c(5, 6) | is_holiday == "True") %>% select(date)
subway_crime_wkday <- right_join(subway_crime, wkday, c("CMPLNT_FR_DT" = "date"))
subway_crime_wkend <- right_join(subway_crime, wkend, c("CMPLNT_FR_DT" = "date"))
crime_group_station_wkday <- subway_crime_wkday %>% group_by(Station_Name_w_Line) %>% summarize(crime_count = n())
crime_group_station_wkend <- subway_crime_wkend %>% group_by(Station_Name_w_Line) %>% summarize(crime_count = n())

station_total_traffic <- subway_total_traffic %>% group_by(Station_Name_w_Line) %>% summarise(total_traffic = sum(total_vol))
station_total_traffic_wkday <- subway_total_traffic_wkday %>% group_by(Station_Name_w_Line) %>% summarise(total_traffic = sum(total_vol))
station_total_traffic_wkend <- subway_total_traffic_wkend %>% group_by(Station_Name_w_Line) %>% summarise(total_traffic = sum(total_vol))

station_traffic_crime <- inner_join(station_total_traffic, crime_group_station)

station_traffic_crime_wkday <- inner_join(station_total_traffic_wkday, crime_group_station_wkday)
station_traffic_crime_wkend <- inner_join(station_total_traffic, crime_group_station_wkend)

fig6.1 <- ggplot(station_traffic_crime, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Crime Count against Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig6.1
fig6.2 <- ggplot(station_traffic_crime_wkday, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Crime Count against Weekday Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig6.2
fig6.3 <- ggplot(station_traffic_crime_wkend, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Crime Count against Weekend Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig6.3
fig6.4 <- ggplot(station_traffic_crime_wkend, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Crime Count against Weekend Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red") + geom_text(label=station_traffic_crime_wkend$Station_Name_w_Line, check_overlap = TRUE, nudge_y = 0.2) 
fig6.4

station_traffic_felony <- inner_join(station_total_traffic, crime_group_station_felony)
station_traffic_misdemeanor <- inner_join(station_total_traffic, crime_group_station_misdemeanor)
station_traffic_violation <- inner_join(station_total_traffic, crime_group_station_violation)

fig6.5 <- ggplot(station_traffic_felony, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Felony Count against Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig6.5
fig6.6 <- ggplot(station_traffic_misdemeanor, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Misdemeanor Count against Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig6.6
fig6.7 <- ggplot(station_traffic_violation, aes(x = total_traffic, y = crime_count)) + geom_point(color='blue') + ggtitle('Violation Count against Human Traffic - 1 point = one station') + xlab("Human Traffic") + ylab("Crime Count") + stat_smooth(method = "lm", col = "red")
fig6.7

```

In the previous set of scatter plots, we were using each point to represent a day aggregating across all subway stations.  For this set of scatter plts, we instead aggregated across time and let each point represent a unique subway station.  The focus here was to investigate if a subway station with higher traffic also suffered from higher crime rate.  

We noticed a general trend of higher crime count for stations with higher human traffic, and this was true regardless of weekday or weekend, or crime type.  There were two outliers with lower traffic but very high crime count (23 ST on line 6 and 125 ST on line 4, 5 & 6), which meant that for these two stations, their higher crime rate could not be well explained by human traffic alone.  Other factors affecting crime rate could be whether or not that neighborhood tended to have higher crime rate.  Also, lower traffic could also work in the reverse, as a station that is more isolated may attract more potential offenders, since their crimes could be more easily committed unseen.

**Time Series of Crime Count across time; superimposed with Subway Human Traffic, Rainfall**

```{r, echo=FALSE}
fig7.1 <- ggplot(crime_count_date, aes(CMPLNT_FR_DT, y=crime_count)) + geom_line() + ggtitle("Crime Count across Time") + xlab("Time") + ylab("Crime Count") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.1

wkday <- unique(wkday)
wkend <- unique(wkend)
crime_count_date_wkday <- right_join(crime_count_date, wkday, c("CMPLNT_FR_DT" = "date"))
crime_count_date_wkend <- right_join(crime_count_date, wkend, c("CMPLNT_FR_DT" = "date"))

fig7.2 <- ggplot(crime_count_date_wkday, aes(CMPLNT_FR_DT, y=crime_count)) + geom_line() + ggtitle("Crime Count across Time (Weekdays)") + xlab("Time") + ylab("Crime Count") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.2

fig7.3 <- ggplot(crime_count_date_wkend, aes(CMPLNT_FR_DT, y=crime_count)) + geom_line() + ggtitle("Crime Count across Time (Weekends)") + xlab("Time") + ylab("Crime Count") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.3

fig7.4 <- ggplot(crime_type_count_date, aes(CMPLNT_FR_DT, y=crime_count, color=LAW_CAT_CD, group=LAW_CAT_CD)) + geom_line() + ggtitle("Crime Count across Time") + xlab("Time") + ylab("Crime Count") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.4

top_crime_stations_list <- crime_group_station_top$Station_Name_w_Line
subway_crime_top_5 <- subway_crime %>% filter(Station_Name_w_Line  %in% top_crime_stations_list)

top_stations_crime_count_date <- subway_crime_top_5 %>% group_by(CMPLNT_FR_DT) %>% summarize(crime_count = n())
top_stations_crime_type_count_date <- subway_crime_top_5 %>% group_by(CMPLNT_FR_DT, LAW_CAT_CD) %>% summarize(crime_count = n())
fig7.5 <- ggplot(top_stations_crime_count_date, aes(CMPLNT_FR_DT, y=crime_count)) + geom_line() + ggtitle("Crime Count across Time for Top 5 Stations") + xlab("Time") + ylab("Crime Count") +  geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.5

fig7.6 <- ggplot(top_stations_crime_type_count_date, aes(CMPLNT_FR_DT, y=crime_count, color=LAW_CAT_CD, group=LAW_CAT_CD)) + geom_line() + ggtitle("Crime Count across Time for Top 5 Stations") + xlab("Time") + ylab("Crime Count") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.6

fig7.7 <- ggplot(subway_total_traffic_wkday_day, aes(date, y=traffic)) + geom_line() + ggtitle("Subway Human Traffic across Time (Weekdays)") + xlab("Time") + ylab("Human Traffic") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.7

fig7.8 <- ggplot(subway_total_traffic_wkend_day, aes(date, y=traffic)) + geom_line() + ggtitle("Subway Human Traffic across Time (Weekends)") + xlab("Time") + ylab("Human Traffic") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.8

subway_total_traffic_day <- subway_total_traffic %>% group_by(date) %>% summarize(traffic = sum(total_vol))

crime_count_traffic_day <- inner_join(subway_total_traffic_day, crime_count_date, c("date" = "CMPLNT_FR_DT"))
crime_count_rainfall_day <- inner_join(rainfall_day, crime_count_date, c("date" = "CMPLNT_FR_DT"))
traffic_rainfall_day <- inner_join(subway_total_traffic_day, rainfall_day)

# Note: for apply function, "2" indicates an operation on col, while 1 indicates an op on row
crime_count_traffic_day_std <- apply(select(crime_count_traffic_day, -date), 2, function(x) 100 * x / x[1])
crime_count_traffic_day_std <- data.frame(crime_count_traffic_day_std)

crime_count_traffic_day_std <- cbind(date = crime_count_traffic_day$date, crime_count_traffic_day_std)
crime_count_traffic_day_std_tidy <- gather(crime_count_traffic_day_std, key = param, value = value, -date)

fig7.9 <- ggplot(crime_count_traffic_day_std_tidy, aes(date, y=value, color=param, group=param)) + geom_line() + ggtitle("Human Traffic & Crime Count across Time") + xlab("Time") + ylab("Human Traffic, Crime Count") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.9

crime_count_rainfall_day_std <- apply(select(crime_count_rainfall_day, -date), 2, function(x) 100 * (x+0.01) / (x[1]+0.01))
crime_count_rainfall_day_std <- data.frame(crime_count_rainfall_day_std)

crime_count_rainfall_day_std <- cbind(date = crime_count_rainfall_day$date, crime_count_rainfall_day_std)
crime_count_rainfall_day_std_tidy <- gather(crime_count_rainfall_day_std, key = param, value = value, -date)

fig7.10 <- ggplot(crime_count_rainfall_day_std_tidy, aes(date, y=value, color=param, group=param)) + geom_line() + ggtitle("Crime Count & Rainfall across Time") + xlab("Time") + ylab("Crime Count, Rainfall") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.10

traffic_rainfall_day_std <- apply(select(traffic_rainfall_day, -date), 2, function(x) 100 * (x+0.01) / (x[1]+0.01))
traffic_rainfall_day_std <- data.frame(traffic_rainfall_day_std)

traffic_rainfall_day_std <- cbind(date = traffic_rainfall_day$date, traffic_rainfall_day_std)
traffic_rainfall_day_std_tidy <- gather(traffic_rainfall_day_std, key = param, value = value, -date)

fig7.11 <- ggplot(traffic_rainfall_day_std_tidy, aes(date, y=value, color=param, group=param)) + geom_line() + ggtitle("Human Traffic & Rainfall across Time") + xlab("Time") + ylab("Human Traffic, Rainfall") + geom_smooth(method = "loess", span = .5, se = FALSE)
fig7.11


```


The above set of time series was meant to explore the trend of crime across time and also to see if it varied in the same direction as the other parameters of human traffic and rainfall.  

Firstly, we could see large fluctuations in crime count across time.  This would not be caused by intra-day patterns because we were already looking at total crime count per day.  Therefore, we looked at the crime count across time only for weekdays and then only for weekends, but the fluctations remained.  It was clear that there were other factors influencing crime count not reflected in our analysis.  

Ignoring the  noise, it seemed like crime rate was lower near the start and end of the year with two prominent peaks between Apr and Jun.

Then, we looked at crime count by crime type.  The picture was even less clear, with the more fluctuations dominanting the graph.

In an attempt to isolate the noise, we looked at the crime count for just the top 5 stations (in terms of highest crime rate).  The peaks between Apr and Jun were still there but there was now a new higher peak in Nov which was not visible previously in the overall chart.  The same peak could be observed when we looked at crime rate by type.

Next, we looked at human traffic and crime count across time and there seemed to be a positive correlation.  This matched our earlier observation with scatter plots. 

We further tried to look at rainfall with human traffic, and rainfall with crime rate, but these two graphs were dominated by large fluctuations in rainfall and were not informative.  It was clear to us that scatter plot offered a better way to visualize relationship between different variables especially if we did not believe that time was a major influencing factor for two parameters (e.g., crime and rainfall) in the same way. 


#3. Description of Data

## 3.1 Crime Data
Crime data was collected from the following website provided by the city of New York:
https://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Historic/qgea-i56i

The original data set comprised 5.58M rows with 24 cols.  The oldest case happened on Jan 1, 1948 and the most recent case in the dataset was on Dec, 31, 2016.

### 3.1.1 Extracting only the relevant rows - 2015 data, Crime in Subway Stations, Manhattan Stations
As the most recent taxi data was for year 2015, we used the year 2015 for crime data to be consistent.   After filtering for only 2015 data, the crime dataset was reduced to 477,031 rows.  

We then filtered the data to include only the crimes committed at subway stations.  This further reduced the number or rows down to 11,228.

Lastly, we filtered for only the stations in Manhattan.  This reduced the number of rows down to 5,367.

### 3.1.2 Include time window and date/time conversion
As the date and time of the data set was in string type, we had to convert it to date and time format in order to filter the data by date and time.  Regarding time, subway human traffic data was organized in 4 hourly interval.  To facilitate analysis of crime data with subway human traffic data, we inserted an additional column in this crime dataset to indicate which 4 hourly  window the crime was commmited in.

### 3.1.3 Remove rows with null lat /long
The crime dataset on its own did not indicate in which subway station the crimes were committed.  Therefore, we wrote a program that assigned a crime location to the closest subway station based on distance calculated using lat/long difference.

For this purpose, it was therefore important that the data had lat/long entries.  However, 170 of the filtered records had null for lat/long.  We removed these records from our analysis.

### 3.1.4 Drop columns
As earlier mentioned, there were 24 columns in the dataset.  Many of these columns were not necessary for our analysis.  For instance, we did not need to know the date and time at which the crime was reported.  We were interested in the date/time at which the crime was *committed* and this information was available in the dataset.  We were also not interested in the codes used for each crime type which were for the police department's internal reference.  We were interested in the type of crime expressed in a language familiar to the lay person.  For that, the dataset conveniently included a column to indicate which of 3 categories the crime committed was in: felony, misdemeanor, violation.  There were also other columns giving a detailed description of the crime but for the purpose of faceting and grouping, we felt that these information was too granular and we did not include them.  Lastly, there were also other columns indicating if the crime was committed in a park, which we did not includ as these were also irrelevant for our analysis.  

In the end, we kept 12 columns the highlights of which included: crime commmited date, time, lat/long, crime category.

We added 6 columns, which as earlier mentioned were columns that we engineered to capture in which 4 hourly window the crime was committed, plus in which subway stations (based on the results of our program to estimate the subway stations based on lat/long proximity)  

## 3.2 Weather Data
Weather data was collected from the following website provided by the National Centers for Environmental Information (NOAA):
https://www.ncdc.noaa.gov/cdo-web/datatools/lcd

There was only one weather station for selection in Manhattan, located at Central Park.  Since Manhattan is relatively small and it is unlikely there to be large variations in weather conditions in different parts of Manhattan, this was sufficient for the purpose of our project.

Unlike the crime data for which we had to download the entire historical set and then filter for 2015 data on our own, we were able to specify the date range within the website before we downloaded the data.
  
The downloaded dataset comprised 13,329 rows and 90 cols.

The weather dataset contained quite a rich set of weather-related data, such as humidity, average wind speed, monthly minimum temperature, hourly sky conditions.  For this dataset, we decided to focus only on rainfall as the weather condition to use when analyzing weather's relationship with crime and subway human traffic data.

### 3.2.1 Include time window and date/time conversion
Similar to what we did for crime data, we converted the date and time from string type to date and time format.  Likewise, we inserted an additional column in this weather dataset to indicate which 4 hourly  window the weather recording was made in.

### 3.2.2 Drop columns
As earlier mentioned, there were 90 columns in the dataset and many of these columns were not related to rainfall and necessary for our analysis.  Eventually, this dataset was transformed into one with only 3 columns, all of which were engineered features.  We explain the feature engineering below.

### 3.2.3 Clean up  rainfall readings
There were several columns representing rainfall, but we focused only on the one giving hourly rainfall (i.e., the highest time resolution).  For hourly rainfall, there were readings with 'T' for value instead of a numerical reading.  There were 1,991 such readings.  'T' represented 'trace' meaning a very small amount of rainfall which were too small for the instrument to provide an accurate reading.  For these readings, we replaced them with 0.

There were also 1,761 hourly rainfall readings with a null value.  Likewise, we replaced them with '0'.

There were hourly readings appended with the character 's' to indicate that these were estimated readings.  For the purpose of our project, we did not have to differentiate between estimated and actual measured rainfall.  Therefore, we removed this information.

Lastly, the hourly rainfall readings were recorded as strings which we could not use for numerical operations.  We converted them to numerical numbers.

### 3.2.4 Aggregate rainfall readings
As consistent with the entire project, we were interested in data up to a 4hourly window resolution.  The hourly rainfall data had to be converted to a similar format to facilitate our analysis.  Therefore, we aggregated the rainfall readings within each 4 hourly window to get the total rainfall measured within that window.

As mentioned earlier, we transformed this weather dataset into 3 columns: they are date, 4hourly time interval and total rainfall (within that 4hourly time interval).

## 3.3 Taxi Data
Taxi data was collected from the following website provided by the city of New York:
https://data.cityofnewyork.us/Transportation/2015-Yellow-Taxi-Trip-Data/ba8s-jw6u

2015 Yellow Taxi Trip data includes all taxi trips completed in yellow taxis in New York from January to June in 2015. This dataset is collected by technology providers authorized under the Taxicab Passenger Enhancement Program and provided to the NYC Taxi and Limousine Commission. This dataset is published on NYC Open Data. The recorded fields are pick-up and drop-off dates and locations, trip distances, payment methods, fares, etc. There are total of 77.1 million rows. Due to the large size of the dataset, we decided to work with trips completed in February, because February has the most average daily trips, which is approximately 443,000 trips per day. 

Analysis of Data Quality
The purpose of using 2015 Yellow Trip data was to map pick-up and drop-off locations to the nearest subway stations in Manhattan and visualize behavioral patterns between subway riders and taxi riders. Only 1.7% of the dataset has missing pick-up and drop-off locations. And this dataset contains all taxi trips in 2015, which guarantees no bias. We understand that February data won’t represent the entire dataset, but for simplicity, we excluded seasonality from our analysis.



